---
output: github_document
always_allow_html: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(tidyverse)
library(vpstheme)
library(geomtextpath)
library(kableExtra)
```

```{r update_sqwiggle, include=FALSE}
source("./r/sqmiggle_2026.R")
```

# Sqwiggle 

## Introduction

<!-- badges: start -->

<!-- badges: end -->

Sqwiggle is a simple [ELO](https://cran.r-project.org/web/packages/fitzRoy/vignettes/elo-ratings-example.html) model for predicting [AFLW](https://www.afl.com.au/aflw) and [AFLM](https://www.afl.com.au)results built in R using the [fitzRoy](https://cran.r-project.org/web/packages/fitzRoy/index.html) and [elo](https://cran.r-project.org/web/packages/elo/index.html) packages. The name is derived from [Squiggle](https://squiggle.com.au/).

## A better ladder

A key motivation for Sqwiggle was to better understand the state of play in the AFLW, which is an uneven competition with a compromised fixture.

Sqwiggle's tipping performance for AFLW 2025 was admirable. Sqwiggle would have won [The Age's expert tips competition](https://www.theage.com.au/sport/afl/aflw-expert-tips-20250814-p5mn1s.html) and finished top ten in [ESPN's footy tips competition](https://www.footytips.com.au/competitions/aflw/ladders?view=ladderScores&season=747&competitionId=706852).

For 2026, I have rebuilt sqwiggle to predict AFLM results. The predicted ladder after opening round is below.


```{r ladder_sqwiggle, echo=FALSE}

combined_ladder |>
  rename_with(\(x) str_replace_all(x,"_|club", " ") |> str_to_sentence()) |> 
  knitr::kable() |> 
   kable_styling()  |> 
   row_spec(1:8, background = bv.crystal) |> 
   row_spec(9:18, background = bv.white)

```


## Team strength

As an ELO model, Sqwiggle's assessment of teams and predictive ability is based on past performance. Sqwiggle learns as each round is played. Sqwiggle accounts for distance travelled when calculating home ground advantage. 

Sqwiggle's estimate of each team's ELO score from the start of the season to after opening round is shown below.

```{r sqwiggle_wiggle, echo=FALSE, out.width = 2000, out.height = 1000}

suppressWarnings(
  ggplot(
    data = sqmiggles_2026 |> mutate(ymin = 1500, xmin = 0),
    aes(
      x = round_number, 
      y = value, 
      xmax = round_number, 
      ymax = value, 
      ymin = ymin, 
      xmin = xmin, 
      colour = team_club_name, 
      group = team_club_name,
      label = team_club_name, 
      fill = team_club_name)
  ) +
  geom_hline(
    yintercept = 1500, 
    colour = bv.charcoal, 
    linetype  = 2) +
  geom_line() +
  geom_ribbon(outline.type = "upper") +
  theme_vps_dh(base_size = 7) +
  facet_wrap(
    facets = vars(team_club_name),
    ncol = 6,
    strip.position = "top") +
  scale_y_continuous(
    limits = c(1240,1660),
    breaks = seq(1200,1700,by = 50),
    name = "Sqwiggle score") +
  scale_x_continuvps(
    breaks = 1L:18L, 
    name = NULL, limits = c(-1.1,last_round+.7)) +
  scale_colour_manual(
   values = unlist(team_colours))+
  scale_fill_manual(
    values = alpha(unlist(team_colours),.5))+
  guides(
    colour = "none", fill = "none") +
  labs(
    caption = "Note: 1500 is the base score in Sqwiggle.")+
  theme(
    axis.line.x = element_blank(),
    strip.placement = "inside")
)
```

## How well did Sqwiggle go at tipping the AFLM in 2026

Keep track of tipping performance here as the season progresses.

```{r summarise_sqwiggle, echo=FALSE}

tipping_results |> 
  group_by(round_round_number) |> 
  summarise(correct = sum(correct), games = n()) |> 
  mutate(cumulative = cumsum(correct)) |> 
  filter(round_round_number <= last_round) |> 
  rename_with(
    \(x) str_replace_all(x,"round_|_|club_name", " ")
    |> str_to_sentence()) |> 
  knitr::kable() |> 
   kable_styling() 

```

Sqwiggles tips for the next round are below.

```{r tipped_by_sqwiggle, echo=FALSE}

tips_2026_m |> 
  filter(round_round_number == (last_round + 1)) |> 
  mutate(home_game_advantage=round(home_game_advantage)) |> 
  rename_with(
    \(x) str_replace_all(x,"round_number|_|club_name", " ")
    |> str_to_sentence()) |> 
  knitr::kable() |> 
  kable_styling()

```
